Random Matrix Theory Tools for the Predictive Analysis of Functional Magnetic Resonance Imaging Examinations

### 2023-04-07 paper/paper/SPIE/response_letter.tex

\title{Cover Letter}
% \begin{center}
% \includegraphics[width=4in]{stfx_logo.png}
% \end{center}

\longcomment{
Specific comments:

% 1. Lines 42-45: I think it's unlikely that most readers of SPIE will be
% familiar with RMT and so it would be a good idea to have a bit more of an
% introduction - what's a random matrix?, what does RMT try to do (e.g.,
% characterize systems that can be described using matrices, statistically,
% rather than deterministically.) Later they talk about what RMT has been applied
% to, but it would be good to orient the reader to the general idea.

%%%%%% 2. Line 44: "independently" should be "independent"

%%%%%%% 3. Line 56: This is a pretty broad statement, because it would seem that it's
%%%%%%% an open question whether all complex systems would lend themselves to RMT-based
%%%%%%% analysis.

%%%%%% 4. Line 100: Do the authors really mean "eliminate" here? Do they instead mean
%%%%%% "mask", given that the information presumably doesn't disappear.

%%%%%% 5. Line 121: "functional connectivity". Do the authors mean "functional
%%%%%% connectivity matrix"?

%%%%%% 6. Line 138: This sentence is a bit garbled. Do the authors mean "often poorly
%%%%%% documented or even missing..."?

%%%%%% 7. Lines 143-145: I don't think these absences raise doubts about these two
%%%%%% things in general, but rather that one has to carefully document the workflows
%%%%%% and make the data and code available to the public.

%%%%%% 8. Line 156: Do the authors mean "allow a reasonable classification of each
%%%%%% scan without resorting to other information". Unclear what they mean by
%%%%%% "entirety".

%%%%%% 9. Table 1: Volumes and n_scans are not defined sufficiently. Are "Volumes" the
%%%%%% number of volumes per run? Is n_scans to total number of runs? Is there one run
%%%%%% per participant? The participant info is in Table 2, but Table 1 should be
%%%%%% independently clear.

%%%%%% 10. Line 180: "learning vs. rest" should be in a different font to be
%%%%%% consistent with the other dataset descriptions.

%%%%%% 11. Line 208: Are the "fractional anisotropy" values mentioned here the input
%%%%%% variables for BET which govern how aggressive BET is and the gradient
%%%%%% parameter?

%%%%%% 12. Section 3.2: While I understand that the authors are trying to make a
%%%%%% general statement that will apply to all features, this section is unclear
%%%%%% because the reader has no idea at this point what features are. Perhaps this
%%%%%% could be expanded.

%%%%%% 13. Line 223: How about "computing time" and "storage space", assuming that is
%%%%%% what is meant.

%%%%%% 14. Line 267: It's not clear that this section should be here, maybe at the end
%%%%%% of the methods section?

%%%%%% 15. Line 292: This is a minor point, but sentences shouldn't start with an
%%%%%% abbreviation (E.g.) better "For example,". This also occurs in line 387.

%%%%%% 16. Line 311: I see what they mean by "voxel time series" because of lines
%%%%%% 317-318, so it might be clearer to move those lines to directly after line 311,
%%%%%% so it is clear without having to read further.

@@

Thank you for your careful consideration and comments on our manuscript
\emph{Random Matrix Theory Tools for the Predictive Analysis of Functional
Magnetic Resonance Imaging Examinations} submitted for publication in SPIE
Journal of Medical Imaging.

In response to reviewer comments, we have made a number of changes to the
manuscript. Minor changes relating to wording or for which an edit is
straightforward (changing or adding a word or two) are identified in
\textcolor{orange}{orange} in the updated \.pdf manuscript (or with
\verb|\textcolor{orange}{text}| in the \.tex file). Other more significant
changes requiring the addition (or movement) of 1-3 sentences are identified
in \textcolor{cyan}{cyan} (\verb|\textcolor{cyan}{text}|).

\begin{description}
\item[Comment 1]: An introduction to the basics of Random Matrix Theory (RMT) has been
added to the introduction, to improve the basic accessibility of RMT to readers
\end{description}

Reviewer comments 2, 4, 5, 6, 10, 11, 13, 14, 15, 16 noted easily-correctable
and uncontroversial issues relating to phrasing, clarity or ordering, and we
have made changes in line with those comments.

Comment 3 interprets our statement as making a broad claim, but the statement was
in fact intended to qualify or limit the scope of applicability of RMT. This has been
made clearer. Likewise, we don't think there is any disagreement between our
intended meaning and the sentiment in comment 7, but have made it clearer that
our doubts are confined to \textit{past} studies (e.g. with poorly documentation,
as mentioned in that paragraph). Of course we don't think poor past studies mean no
future for RMT in general.

Reviewer comments 8, 9 noted some issues with vagueness involving descriptions
of the fMRI scans and runs in the text and tables. We hope these descriptions are now
unambiguous.

In response to comment 12, while we did, and still do take as given that our
reviewer and a reader of SPIE:JMI knows what features and feature extraction
are, we can see how at this point in the paper, it may not be clear
\textit{why} we are extracting features. We have added a short paragraph here
that we hope resolves this, and makes it clear we are comparing the predictive
utility of different features / feature extraction methods (and that "feature"
has the usual meaning in this context).

### 2023-04-07 paper/paper/SPIE/article.tex
wip

At its most rudimentary, RMT describes the expected behavior of the
eigenvalues—also often called levels—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. A random matrix is,
simplifying somewhat, a matrix in which the entries are drawn from
\textcolor{orange}{independent} and identically distributed (\textit{iid})
distributions. Though first developed to describe the fluctuation of nuclei
energy levels in quantum physics \cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}, RMT has been shown to have extremely
broad potential. In small-scale physical systems, RMT universalities have been
observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,
mehtaRandomMatrices2004, brodyRandommatrixPhysicsSpectrum1981,
beenakkerRandommatrixTheoryQuantum1997,
bohigasHigherOrderCorrelationsSpectra1985, wintgenLevelStatisticsQuantized1988,
pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has
been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002}, social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids\cite{sastrySpectralStatisticsInstantaneous2001,
matharooSpectralStatisticsQuenched2009} and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,
matharooVibrationalSpectraAmorphous2005,
matharooUniversalityVibrationalSpectra2008}. Within biological systems, RMT has
also been used to successfully model aspects of  amino acid functional
relationships \cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in
epileptic seizures \cite{osorioPhasesynchronizationRandommatrixBased2011}, and
in protein-protein interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep neural
networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

RMT likely has the most real-world explanatory potential when dealing with a
complex system of many (hundreds or more\cite{mehtaRandomMatrices2004})
interacting components. If such a system has a matrix representation, and the
eigenvalues of this representation have a distribution similar to one predicted
by RMT, it suggests the system is either highly random or chaotic. By contrast,
if the observed spectra deviate significantly from RMT-predicted spectra, this
suggests otherwise. A number of studies have used RMT to make such
interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

In addition to the flexibility in the implementation and application of RMT to
empirical data, there is also flexibility introduced by analytic choices made
in the complex preprocessing pipelines of
fMRI\cite{parkerBenefitSliceTiming2019}. These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b}, in combination
with the absence of reproducible code and data availability, and predominance
of descriptive and explanatory approaches, raise doubts about the basic
robustness and practical utility of RMT-based functional connectivity analyses.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro}
platform\cite{markiewiczOpenNeuroResourceSharing2021}. Selection criteria was
somewhat subjective, but we required each dataset to (1) comprise 10 or more
human subjects, (2) have all fMRI scans have the same spatial and temporal
resolutions, (3) have fMRI images that can be split into various classes for a
number of binary classification tasks, and (4), allow reasonably classifying
each scan in its entirety (i.e. neither task nor event timings are needed to
determine class membership, nor do any scan volumes need to be removed to make
the scan classifiable). This yielded seven datasets total.

\begin{table}[h!]
\caption{
    \label{tab:data-dimensions}
    Included fMRI dataset details.
    Name = Identifier for paper.
    Dimensions listed as M x N x P, indicate P axial slices each with dimensions M x N.
    TR = Time of Repetition (seconds).
    n\_scans = Total number of images in dataset.
}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

@@

For each preprocessing degree and fMRI image, features were extracted from all
\(N\) non-constant brain voxels, where \(N\) may differ for each image. After
this process, each scan yields an \(N \times t\) matrix \(\mathbf{M}\), where
\(t\) is the number of volumes acquired. All features extracted summarize
this matrix \(\mathbf{M}\).

@@

\textit{Software Release}. Because the computation of these statistics is
non-trivial, and to make the unfolding procedure more transparent and
reproducible, we developed and release a separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances. The library also makes available
various other functions useful for empirical RMT analyses, such as unfolding
and trimming functions, and plotting facilities for classic RMT ensembles.

@@

>>>>

\color{cyan}

At its most rudimentary, the basic insight of Random Matrix Theory (RMT) is
that matrices can be described by their eigenvalues, and that
large matrices (or large collections of matrices) can be
well-described by the statistical behaviour of those eigenvalues. Such a
description is obtained by assuming that the large matrix (or collection
of matrices) is sampled from a \textit{class} or \textit{ensemble} of random
matrices.

Simplifying somewhat, a random matrix is a matrix that has entries that can be
well-described as being drawn from a particular independent and identically
distributed (\textit{iid}) distribution, or, equivalently, is a matrix where
each entry is itself a random variable. A \textit{class} or \textit{ensemble}
of random matrices is generally defined by the shape of the \textit{iid}
distribution involved. For example, the Gaussian Orthogonal Ensemble (GOE) is a
\textit{class} or \textit{ensemble} of random matrices that comprises
orthogonal matrices with each entry being sampled from a standard Gaussian, and
the Marchenko-Pastur distribution describes the limiting behaviour of the
singular values of \textit{iid} rectangular random matrices where the
\textit{iid} distribution is arbitrary\cite{mehtaRandomMatrices2004}. Of
course, given a single particular matrix, one can only estimate a
\textiit{probability} that it is a member of a certain class, since certain RMT
classes can yield arbitrary matrices (albeit with extremely low probability).

and eigenvalue distributions of GOE random matrices have a
number of well-understood properties\cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}. Even more generally,

Specifically, RMT finds that often, in the infinite limit, that the expected
distribution of the eigenvalues of a number of classes of random matrices can
be described quite precisely.\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. Although these
assumptions are quite specific and restrictive, nevertheless, RMT-predicted
eigenvalue distributions have appeared in diverse natural phenomena.

\color{black}

\textcolor{cyan}{Thus, while} first developed to describe the fluctuation of nuclei
energy levels in quantum physics \cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}, RMT has been shown to have extremely
broad potential. In small-scale physical systems, RMT universalities have been
observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,
mehtaRandomMatrices2004, brodyRandommatrixPhysicsSpectrum1981,
beenakkerRandommatrixTheoryQuantum1997,
bohigasHigherOrderCorrelationsSpectra1985, wintgenLevelStatisticsQuantized1988,
pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has
been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002}, social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids\cite{sastrySpectralStatisticsInstantaneous2001,
matharooSpectralStatisticsQuenched2009} and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,
matharooVibrationalSpectraAmorphous2005,
matharooUniversalityVibrationalSpectra2008}. Within biological systems, RMT has
also been used to successfully model aspects of  amino acid functional
relationships \cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in
epileptic seizures \cite{osorioPhasesynchronizationRandommatrixBased2011}, and
in protein-protein interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep neural
networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

\textcolor{orange}{When or if RMT has} real-world explanatory potential,
\textcolor{orange}{this will most likely be} when dealing with a complex system
of many (hundreds or more\cite{mehtaRandomMatrices2004}) interacting
components, \textcolor{orange}{rather than a system that is too small for
statistical regularities ro reliably surface}. If such a system has a matrix
representation, and the eigenvalues of this representation have a distribution
similar to one predicted by RMT, it suggests the system is either highly random
or chaotic. By contrast, if the observed spectra deviate significantly from
RMT-predicted spectra, this
suggests otherwise. A number of studies have used RMT to make such
interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

In addition to the flexibility in the implementation and application of RMT to
empirical data, there is also flexibility introduced by analytic choices made
in the complex preprocessing pipelines of
fMRI\cite{parkerBenefitSliceTiming2019}. These "researcher degrees of
freedom"\cite{simmonsFalsePositivePsychologyUndisclosed2011b}, in combination
with the absence of reproducible code and data availability, and predominance
of descriptive and explanatory approaches, \textcolor{orange}{may} raise doubts about the basic
robustness and practical utility of \textcolor{orange}{past} RMT-based functional connectivity analyses.

@@

We selected fMRI data publicly available on the
\href{https://openneuro.org/}{OpenNeuro}
platform\cite{markiewiczOpenNeuroResourceSharing2021}. Selection criteria was
somewhat subjective, but we required each dataset to (1) comprise 10 or more
human subjects, (2) have all fMRI \textcolor{orange}{runs} have the same
spatial and temporal resolutions, (3) have fMRI images that can be split into
various classes for a number of binary classification tasks, and (4), allow
reasonably classifying each run \textcolor{orange}{using all of the run 4D
voxel data only.} \textcolor{cyan}{That is, for this last point, classification
of a run should be reasonable in the absence of run-specific task or event
timings or details, and should not require using only some subset of the total
set of 3D volumes (e.g. those associated only with some task or event
timings)}. This yielded seven datasets total.

\begin{table}[h!]
\caption{
    \label{tab:data-dimensions}
    Included fMRI dataset details.
    Name = Identifier for paper.
    Dimensions listed as M x N x P, indicate P axial slices each with dimensions M x N.
    TR = Time of Repetition (seconds).
    \textcolor{orange}{Volumes = Number of 3D volumes per run.}
    n\_scans = Total number of \textcolor{orange}{4D images in dataset (number of subjects times number of runs per subject)}.
}
\small
\centering
\begin{tabular}{ l c c c c c c }
\hline
\textbf{Name}    & \textbf{Dimensions}  & \textbf{Voxel size (mm)} & \textbf{TR} & \textbf{Volumes} & \textbf{n\_scans} \\
\hline
Aging     & 74 × 74 × 32   & 3.0 × 3.0 × 4.0 & 2.0  & 300 & 62  \\
Bilingual    & 100 × 100 × 72 & 1.8 × 1.8 × 1.8 & 0.88 & 823 & 90  \\
Depress   & 112 × 112 × 25 & 2.0 × 2.0 × 5.0 & 2.5  & 100 & 72  \\
Learn     & 64 × 64 × 36   & 3.0 × 3.0 × 3.0 & 2.0  & 195 & 432 \\
Osteo     & 64 × 64 × 36   & 3.4 × 3.4 × 3.0 & 2.5  & 300 & 74  \\
Park      & 80 × 80 × 43   & 3.0 × 3.0 × 3.0 & 2.4  & 149 & 552 \\
Attention & 128 × 128 × 70 & 1.5 × 1.5 × 1.5 & 3.0  & 300 & 90  \\
\hline
\end{tabular}
\end{table}

@@

\textcolor{cyan}{When dealing with large inputs, any predictive analysis must
reduce that input into predictively-useful features. Any method that reduces a
larger set of data is potentially valid as a feature extraction method, and RMT,
which summarizes matrix data by summarizing the eigenvalue distributions, may
may be useful for feature extraction. However, different feature
extraction methods may have different advantages and disadvantages in terms of
the computational complexity, amount of summary or selection involved, and in
interpretability. RMT is both mathematically and computationally complex, and
thus RMT-based features should be compared to alternate features that may be
easier to understand or compute.}

Thus, for each preprocessing degree and fMRI image, features were extracted
from all \(N\) non-constant brain voxels, where \(N\) may differ for each
image. After this process, each scan yields an \(N \times t\) matrix
\(\mathbf{M}\), where \(t\) is the number of volumes acquired. All features
extracted summarize this matrix \(\mathbf{M}\).

@@

@@

\subsection{Software Release}

\textcolor{cyan}{Because the computation of these statistics is
non-trivial, and to make the unfolding procedure more transparent and
reproducible, we developed and release a separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances. The library also makes available
various other functions useful for empirical RMT analyses, such as unfolding
and trimming functions, and plotting facilities for classic RMT ensembles.}
